Tactic Script Optimisation for Aesop

### 2024-09-17 paper.tex
Banish filenames into footnotes

The above description omits some special cases, e.g.\ local definitions and delayed-assigned metavariables.
See the implementation in file \texttt{Aesop/Util/EqualUpToIds.lean} for the full details.

@@

However, in practice, \tac{unfold} actually produces the correct result in many cases.
To take advantage of this observation, we let each step contain not just one but multiple alternative tactics.
Of these, only the last must produce the right result; the others may be variants that are more idiomatic but may produce the wrong result.
The \emph{local dynamic optimisation} step (implemented in \texttt{Aesop/Script/Step.lean}) then selects the first tactic that does, in fact, produce the right result.

@@

% TODO check all filenames
The static reordering algorithm is shown as Algorithm~\ref{alg:static-reordering} and implemented in \texttt{Aesop/Script/StructureStatic.lean}.
Its main function, \textsc{reorder-static}, first sets up $\mathit{tacticState}$.
This variable contains the goals of the current tactic state, so we initialise it with the pre-goal of the first step.
(For the purposes of static reordering, metavariable assignments and declarations are not essential, so $\mathit{tacticState}$ does not track the current metavariable context.)
The remainder of the algorithm iteratively selects steps and applies them to the tactic state (simulating the effect of the step's tactic) until there are no goals left.

@@

To overcome the inherent limitations of static reordering, I propose a dynamic variant.
This variant is based on the observation that even if reordering in the presence of metavariables is not always possible, it is still possible often enough to be worth trying.
The algorithm, shown as Algorithm~\ref{alg:dynamic-reordering} and implemented in \texttt{Aesop/Script/StructureDynamic.lean}, therefore reorders optimistically.

@@

Once the script is ordered, it needs to be transformed into a \emph{structured script} (defined in \texttt{Aesop/Script/SScript.lean}).
This is a tree with three different kinds of nodes:
\begin{itemize}
  \item The $\mathrm{empty}$ node with no data and no children.
  \item Nodes of the form $\mathrm{on}(p, s, t)$, where $p$ is a natural number, $s$ is a step and $t$ is a child structured script.
  \item Nodes of the form $\mathrm{fas}(p, h, t)$, where $p$ is again a natural number and $h$ and $t$ are children.
\end{itemize}

@@

Transforming an ordered sequence of steps into a structured script (implemented in \texttt{Aesop/Script/UScriptToSScript.lean}) is conceptually straightforward, but not entirely trivial.
As a start, it is convenient to view the sequence of steps as a \emph{step tree}.
This is a tree with two types of nodes:
\begin{itemize}
  \item An \emph{empty node} denotes an empty tree.
  \item An \emph{inner node} contains a step and an index (a natural number) and has a finite number of children.
\end{itemize}
We construct the step tree for a sequence of steps $s₁, \dots, sₙ$ as follows.
First, construct an inner node for step $s₁$.
Its children are inner nodes for those steps $s_{i₁}, \dots, s_{iₖ}$ whose pre-goals are the post-goals of $s₁$.
In other words, the children of a step are those steps that are applied to the goals generated by the step.
Applying this procedure recursively yields a tree of nodes.
The empty node is only used if the sequence of steps is empty.

@@

In fact, this treatment of the tactic state is slightly simplified.
When generating the $h$ script, we restrict the tactic state to the single goal $g$.
However, this does not mean that $h$ only acts on $g$ and its descendants: steps can also solve goals that are not present in the tactic state.
This happens when a goal appears as a metavariable in one of the goals that are present in the tactic state and a step assigns this metavariable.
The tactic state must therefore track such \enquote{invisible} goals as well and after generating $h$, we must remove any solved goals from the tactic state $ts$.
See the implementation in \texttt{Aesop/Script/TacticState.lean}.

>>>>

The above description omits some special cases, e.g.\ local definitions and delayed-assigned metavariables.
See the implementation\footnote{\texttt{Aesop/Util/EqualUpToIds.lean}} for the full details.

@@

However, in practice, \tac{unfold} actually produces the correct result in many cases.
To take advantage of this observation, we let each step contain not just one but multiple alternative tactics.
Of these, only the last must produce the right result; the others may be variants that are more idiomatic but may produce the wrong result.
The \emph{local dynamic optimisation} step\footnote{Implemented in \texttt{Aesop/Script/Step.lean}.} then selects the first tactic that does, in fact, produce the right result.

@@

% TODO check all filenames
The static reordering algorithm\footnote{Implemented in \texttt{Aesop/Script/StructureStatic.lean}.} is shown as Algorithm~\ref{alg:static-reordering}.
Its main function, \textsc{reorder-static}, first sets up $\mathit{tacticState}$.
This variable contains the goals of the current tactic state, so we initialise it with the pre-goal of the first step.
(For the purposes of static reordering, metavariable assignments and declarations are not essential, so $\mathit{tacticState}$ does not track the current metavariable context.)
The remainder of the algorithm iteratively selects steps and applies them to the tactic state (simulating the effect of the step's tactic) until there are no goals left.

@@

To overcome the inherent limitations of static reordering, I propose a dynamic variant.
This variant is based on the observation that even if reordering in the presence of metavariables is not always possible, it is still possible often enough to be worth trying.
The algorithm\footnote{Implemented in \texttt{Aesop/Script/StructureDynamic.lean}.}, shown as Algorithm~\ref{alg:dynamic-reordering}, therefore reorders optimistically.

@@

Once the script is ordered, it needs to be transformed into a \emph{structured script}.\footnote{Implemented in \texttt{Aesop/Script/SScript.lean} and \texttt{Aesop/\allowbreak Script/\allowbreak UScriptToSScript.lean}.}
This is a tree with three different kinds of nodes:
\begin{itemize}
  \item The $\mathrm{empty}$ node with no data and no children.
  \item Nodes of the form $\mathrm{on}(p, s, t)$, where $p$ is a natural number, $s$ is a step and $t$ is a child structured script.
  \item Nodes of the form $\mathrm{fas}(p, h, t)$, where $p$ is again a natural number and $h$ and $t$ are children.
\end{itemize}

@@

Transforming an ordered sequence of steps into a structured script is conceptually straightforward, but not entirely trivial.
As a start, it is convenient to view the sequence of steps as a \emph{step tree}.
This is a tree with two types of nodes:
\begin{itemize}
  \item An \emph{empty node} denotes an empty tree.
  \item An \emph{inner node} contains a step and an index (a natural number) and has a finite number of children.
\end{itemize}
We construct the step tree for a sequence of steps $s₁, \dots, sₙ$ as follows.
First, construct an inner node for step $s₁$.
Its children are inner nodes for those steps $s_{i₁}, \dots, s_{iₖ}$ whose pre-goals are the post-goals of $s₁$.
In other words, the children of a step are those steps that are applied to the goals generated by the step.
Applying this procedure recursively yields a tree of nodes.
The empty node is only used if the sequence of steps is empty.

@@

In fact, this treatment of the tactic state is slightly simplified.
When generating the $h$ script, we restrict the tactic state to the single goal $g$.
However, this does not mean that $h$ only acts on $g$ and its descendants: steps can also solve goals that are not present in the tactic state.
This happens when a goal appears as a metavariable in one of the goals that are present in the tactic state and a step assigns this metavariable.
The tactic state must therefore track such \enquote{invisible} goals as well and after generating $h$, we must remove any solved goals from the tactic state $ts$.%
\footnote{See the implementation in \texttt{Aesop/Script/TacticState.lean}.}
